Concentration of Maxima and Fundamental Limits in High-Dimensional Testing and Inference

### 2021-01-08 1.intro/introduction.tex
shortened version
* Implemented my corrections up to Ch.3.1

* Moved the proofs of Ch 3 and the sections on non-AGG models from Ch 4 to the Appendix -- which will become the online supplement.

* We need to cut just 5 pages now...  This can happen by re-distributing some proofs in the text (instead of having them in separate sections) and moving some more proofs from Ch.6 to the Appendix.

* Overall, I don't like the idea to eliminate the proofs from Ch. 3.   I also like having the non-AGG models in the main text, too.  I will think if we can shorten the manuscript in alternative ways....

  O

The proliferation of information technology has enabled us to collect and consume huge volumes of data at unprecedented speeds and at very low costs.
This fast and cheap access to data gave rise to fundamentally new ways of pursuing scientific questions.
In contrast with the traditional hypothesis--experiment--analysis cycle where data are collected from the experiments, abundant information are often available before specific questions are even formulated. 
Such information can be used for not just evaluating hypotheses, but also for \emph{generating}, and \emph{selecting} the hypotheses to pursue. 
As a result, {multiple testing} --- where a large number of hypotheses are formulated and screened for their plausibility simultaneously --- has become a staple of modern data-driven studies.

An archetypal example of multiple testing problems is genetic association studies \citep{bush2012genome}.
In genetic association studies, scientists test hypotheses relating each of the hundreds of thousands of genetic marker locations to phenotypic traits of interest;
the goal here is to select the set of most promising genetic markers for subsequent investigation.
Another example of multiple testing problems arise in cybersecurity, where millions of IP addresses are monitored in real time.
In this engineering application, statistics are collected and tests are performed for each IP address, in an attempt to locate the IP addresses with anomalous network activities, so that and malicious traffic and volumetric attacks can be filtered to protect end users of network services \citep{kallitsis2016amon}.

@@

Consider the canonical signal-plus-noise model where the observation $x$ is a high-dimensional vector in $\R^p$,
\begin{equation} \label{eq:model-additive}
    x(i) = \mu(i) + \epsilon(i), \quad i=1,\ldots,p.
\end{equation}
The signal, $\mu = (\mu(i))_{i=1}^p$, is a vector with $s$ non-zero components supported on the set $S=\{i:\mu(i)\neq 0\}$; the second term $\epsilon$ is a random error vector. 
The goal of high-dimensional statistics is usually two-fold: 
\begin{enumerate}
    \item To detect the presence of non-zero components in $\mu$. That is, to test the global hypothesis $\mu = 0$, which we call the \emph{detection} problem, and
    \item To estimate the support set $S$, which we call the \emph{support recovery} problem.
\end{enumerate}
To illustrate, in the engineering application of cybersecurity, Internet service providers (ISP) routinely collect statistics of network traffic to determine if there are abnormal surges or blackouts.
The data vector $x$ could represent, for example, incoming traffic volumes to each server node that the ISP maintains.
In this case, the vector $\mu$ would be the average traffic volumes when under normal operating conditions, and $\epsilon$'s the fluctuations around these normal levels of traffic.
% The ISPs collect, for example, incoming traffic volumes to each server to determine if there are abnormal surges or blackouts.
The signal detection problem in this context is then equivalent to determining if there are \emph{any} anomalies among all servers, and the support recovery problem is equivalent to \emph{identifying} the servers with anomalies.
Similar questions of signal detection and support recovery are pursued in large-scale microarray experiments, %\citep{dudoit2003multiple}, 
brain imaging and fMRI analysis, %\citep{nichols2003controlling}, 
and numerous other applications.

A common theme in such applications is that the errors are correlated, and that the signal vectors are believed to be sparse: the number of non-zero components in $\mu$ is small compared to the number of test performed --- in the cybersecurity example, while monitoring is performed over a large number of servers, it is often believed that only very few servers will be experiencing problems at any time.
Under such sparsity assumptions, it is natural to ask if and when one can reliably {(1)} detect the signals, and {(2)} recover the support set $S$.
% Further, one would like to characterize the statistical procedures that achieve efficient detection and support recovery, whenever such goals become possible to achieve.
In this text, we explore both the \emph{detection} and the \emph{support recovery} problems.  More precisely, we are interested in the 
theoretical feasibility of both problems, and seek minimal conditions under which these problems can be consistently solved in large 
dimensions.

@@

Model \eqref{eq:model-additive} is simple yet ubiquitous.
Consider the linear model
\begin{equation*}
 Y = X\mu + \xi,
\end{equation*}
where $\mu$ is a $p$-dimensional vector of regression coefficients of interest to be inferred from observations of $X$ and $Y$.
If the design matrix $X$ is of full column rank, then the ordinary least squares (OLS) estimator of $\mu$ can be formed 
\begin{equation} \label{eq:OLS}
    \widehat{\mu} = \left(X'X\right)^{-1}X'Y = \mu + \epsilon,
\end{equation}
where $\epsilon := (X'X)^{-1}X'\xi$. \stilian{R2, Typo 2}{I see no problems here except that we are not in the high-dimensional regime $p\gg n$. If we were to address that, we need to expand the discussion. 1) Talk about sparsity of $\mu$ 2) Introduce the assumption that if ${\rm supp}(\mu) = r <n$, then every $r$ of the $p$ columns of 
$X$ must be linearly independent -- is there a name for such a condition in the literature? We know it happens with probability one if $X$ has independent, say Gaussian, entries.  3) Then, it seems clear that the support of $\mu$ will be identifiable if we replace the inverse with a Moore-Penrose inverse. BTW, an intriguing question: Given a design matrix $X$, consider the quantity:
$$
c_r(X) :=\min_{w\subset \{1,\cdots,p\},\ |w| = r} {\| (X_w' X_w)^{-1}\| \over \| X_w'X_w\| },
$$
where the minimum is taken over all subsets of indices $w$ with cardinality $r<n$ and $X_w:= (x_{i,j})_{j\in w}$ is the sub-matrix of $X$ obtained
by selecting the columns indexed by the elements in $w\subset\{1,\cdots,p\}$.  Intuitively, $c_r(X)$ gives us the overall difficulty of identifying a 
sparse signal $\mu$ with support of size $r$. \fbox{Going off on a tangent here...}}
Hence we recover the generic problem \eqref{eq:model-additive}. 
Signal detection is therefore equivalent to the problem of testing the global null model, and support recovery problem corresponds to the fundamental problem of variable selection.

Note that the components in the observation $x$ (and equivalently, the noise $\epsilon$) in \eqref{eq:model-additive} need not be independent. 
In the linear regression example, even when the components of the noise term $\xi$ are independent, those of the OLS estimator \eqref{eq:OLS} need not be, unless we have an orthogonal design.
Indeed, in practice, independence is the exception rather than the rule.
Therefore a general theory of feasibility must address the role of the error dependence structures in such testing problems, and identify practical procedures that attain the performance limits in independent as well as dependent cases, as soon as the problems become theoretically feasible.

>>>>

The proliferation of information technology has enabled us to collect and consume huge volumes of data at unprecedented speeds and at very low costs.
This fast and cheap access to data gave rise to fundamentally new ways of pursuing scientific questions.
In contrast with the traditional hypothesis--experiment--analysis cycle where data are collected from the experiments, \stilian{nowadays abundant data}{abundant information} are often available before specific questions are even formulated. 
Such \stilian{data}{information} can be used for not just evaluating hypotheses, but also for \emph{generating}, and \emph{selecting} the hypotheses to pursue. 
As a result, {multiple testing} --- where a large number of hypotheses are formulated and screened for their plausibility simultaneously --- has become a staple of modern data-driven studies.

An archetypal example of multiple testing problems is genetic association studies \citep{bush2012genome}.
In genetic association studies, scientists test hypotheses relating each of the hundreds of thousands of genetic marker locations to phenotypic traits of interest;
the goal here is to select the set of most promising genetic markers for subsequent investigation.
Another example of multiple testing problems arise in cybersecurity, where millions of IP addresses are monitored in real time.
In this engineering application, statistics are collected and tests are performed for each IP address, in an attempt to locate the IP addresses with anomalous network activities. Thus, malicious traffic and volumetric attacks can be filtered to protect end users of network services \citep{kallitsis2016amon}.

@@

Consider the canonical signal-plus-noise model where the observation $x$ is a high-dimensional vector in $\R^p$,
\begin{equation} \label{eq:model-additive}
    x(i) = \mu(i) + \epsilon(i), \quad i=1,\ldots,p.
\end{equation}
The signal, $\mu = (\mu(i))_{i=1}^p$, is a vector with $s$ non-zero components supported on the set $S=\{i:\mu(i)\neq 0\}$; the second term $\epsilon$ is a random error vector. 
The goal of high-dimensional statistics is usually two-fold: \\

\begin{enumerate}
\item [{\rm I.}] (signal detection or {\em testing}) To detect the presence of non-zero components in $\mu$. That is, to test the global hypothesis $\mu = 0$, which we call the \emph{detection} problem, and\\

\item[{\rm II.}] (support estimation or {\em inference})
 To estimate the support set $S$, which we call the \emph{support recovery} problem.
\end{enumerate}
\medskip
To illustrate, in the engineering application of cybersecurity, Internet service providers (ISP) routinely monitor a large number of network traffic streams to determine 
if there are abnormal surges, blackouts, or other types of anomalies.  The data vector $x$ could represent, for example, incoming traffic volumes to each server 
node, internet protocol (IP) address, or port that the ISP monitors.  In this case, the vector $\mu$ represents the average traffic volumes in each of the streams 
under normal operating conditions, and $\epsilon$'s -- the fluctuations around these normal levels of traffic.  
% The ISPs collect, for example, incoming traffic volumes to each server to determine if there are abnormal surges or blackouts.
The signal detection problem in this context is then equivalent to determining if there are \emph{any} anomalies among all data streams, and the support recovery problem is equivalent to \emph{identifying} the streams experiencing anomalies.  
%The dimension of $x$ (number of streams) is typically very large, while the non-trivial 
%signal is sparse (relatively few data streams experience anomalies).
Similar questions of signal detection and support recovery are pursued in large-scale microarray experiments \citep{dudoit2003multiple},
brain imaging and fMRI analysis \citep{nichols2003controlling}, and numerous other anomaly detection applications.

A common theme in such applications is that the errors are {\em correlated}, and that the signal vectors are believed to be {\em sparse}: the number of non-zero
 (or large) components in $\mu$ is small compared to the number of tests performed.  In the cybersecurity context, while a very large number of data streams are 
 monitored, typically only just a few of them will be experiencing problems at any time, barring large-scale outages or distributed denial of service attacks.
Under such sparsity assumptions, it is natural to ask if and when one can reliably {(1)} detect the signals, and {(2)} recover the support set $S$.
% Further, one would like to characterize the statistical procedures that achieve efficient detection and support recovery, whenever such goals become possible to achieve.
In this text, we explore both the \emph{detection} and the \emph{support recovery} problems.  More precisely, we are interested in the 
theoretical feasibility of both problems, and seek minimal conditions under which these problems can be consistently solved in large 
dimensions.

@@

Model \eqref{eq:model-additive} is simple yet ubiquitous.
Consider the linear model
\begin{equation*}
 Y = X\mu + \xi,
\end{equation*}
where $\mu$ is a $p$-dimensional vector of regression coefficients of interest to be inferred from observations of $X$ and $Y$.
If the design matrix $X$ is of full column rank, then the ordinary least squares (OLS) estimator of $\mu$ can be formed 
\begin{equation} \label{eq:OLS}
    \widehat{\mu} = \left(X'X\right)^{-1}X'Y = \mu + \epsilon,
\end{equation}
where $\epsilon := (X'X)^{-1}X'\xi$. \stilian{R2, Typo 2 -- \fbox{DISCUSS} }{I see no problems here except that we are not in the high-dimensional regime $p\gg n$. If we were to address that, we need to expand the discussion. 1) Talk about sparsity of $\mu$ 2) Introduce the assumption that if ${\rm supp}(\mu) = r <n$, then every $r$ of the $p$ columns of 
$X$ must be linearly independent -- is there a name for such a condition in the literature? We know it happens with probability one if $X$ has independent, say Gaussian, entries.  3) Then, it seems clear that the support of $\mu$ will be identifiable if we replace the inverse with a Moore-Penrose inverse. BTW, an intriguing question: Given a design matrix $X$, consider the quantity:
$$
c_r(X) :=\min_{w\subset \{1,\cdots,p\},\ |w| = r} {\| (X_w' X_w)^{-1}\| \over \| X_w'X_w\| },
$$
where the minimum is taken over all subsets of indices $w$ with cardinality $r<n$ and $X_w:= (x_{i,j})_{j\in w}$ is the sub-matrix of $X$ obtained
by selecting the columns indexed by the elements in $w\subset\{1,\cdots,p\}$.  Intuitively, $c_r(X)$ gives us the overall difficulty of identifying a 
sparse signal $\mu$ with support of size $r$. \fbox{Going off on a tangent here...}}
Hence we recover the generic problem \eqref{eq:model-additive}. 
Signal detection is therefore equivalent to the problem of testing the global null model, and support recovery problem corresponds to the fundamental problem of variable selection.

Note that the components of the observation vector $x$ (and equivalently, the noise $\epsilon$) in \eqref{eq:model-additive} need not be independent. 
In the linear regression example, even when the components of the noise term $\xi$ are independent, those of the OLS estimator \eqref{eq:OLS} need not be, 
unless we have an orthogonal design. Indeed, in practice, independence is the exception rather than the rule.
Therefore, a general theory of feasibility must address the role of the {\em error dependence} structure in such testing and support estimation 
problems.  It is also important to identify practical and/or optimal procedures that attain the performance limits in independent as well as dependent cases, as soon 
as the problems become theoretically feasible.  We address both themes in this text.

### 2021-01-08 2.background/2.AGG.tex
shortened version
* Implemented my corrections up to Ch.3.1

* Moved the proofs of Ch 3 and the sections on non-AGG models from Ch 4 to the Appendix -- which will become the online supplement.

* We need to cut just 5 pages now...  This can happen by re-distributing some proofs in the text (instead of having them in separate sections) and moving some more proofs from Ch.6 to the Appendix.

* Overall, I don't like the idea to eliminate the proofs from Ch. 3.   I also like having the non-AGG models in the main text, too.  I will think if we can shorten the manuscript in alternative ways....

  O

We introduce a fairly general class of distributions known as \ac{AGG} in this section, and state some properties about the tails of the 
\index{AGG}{AGG} distributions.

@@

The AGG models include, for example, the Gaussian distribution ($\nu = 2$), and the Laplace distribution ($\nu = 1$) as special cases. 
Since the requirement is only placed on the tail behavior, this class encompasses a large variety of light-tailed models, and is commonly used in the literature on high-dimensional testing \citep{cai2007estimation, arias2017distribution}.

@@

% \begin{proposition}[$(1 - 1/p)$-th quantile] \label{prop:quantile}
% Let 
% \begin{equation} \label{eq:quantiles-Appendix}
% u_p = F^{\leftarrow}(1-1/p),    
% \end{equation}
% then,
% \begin{equation}
% \lim_{p\to\infty} \frac{u_p}{\left(\nu\log{p}\right)^{1/\nu}} = 1.
% \end{equation}
% \end{proposition}
% 
\begin{proof}[Proof of Proposition \ref{prop:quantile}]
By definition of AGG, for any $\epsilon>0$, there is a constant $C(\epsilon)$ such that for all $x\ge C$, we have
$$
-\frac{1}{\nu}x^\nu(1+\epsilon) \le \log{\overline{F}(x)} \le -\frac{1}{\nu}x^\nu(1-\epsilon).
$$
Therefore, for all $x < x_l := \left((1+\epsilon)^{-1}\nu\log{p}\right)^{1/\nu}$, we have
\begin{equation} \label{eq:AGG-quantiles-proof-1}
    -\log{p} = -\frac{1}{\nu}x_l^\nu(1+\epsilon) \le \log{\overline{F}(x_l)} \le \log{\overline{F}(x)},
\end{equation}
and for all $x > x_u := \left((1-\epsilon)^{-1}\nu\log{p}\right)^{1/\nu}$, we have
\begin{equation} \label{eq:AGG-quantiles-proof-2}
    \log{\overline{F}(x)} \le \log{\overline{F}(x_u)} \le -\frac{1}{\nu}x_u^\nu(1-\epsilon) = -\log{p}.
\end{equation}
By definition of generalized inverse,
\begin{equation*}
    u_p := F^\leftarrow(1-1/p) = \inf\{x:\overline{F}(x)\le 1/p\} = \inf\{x:\log{\overline{F}(x)} \le -\log{p}\}.
\end{equation*}
We know from relations \eqref{eq:AGG-quantiles-proof-1} and \eqref{eq:AGG-quantiles-proof-2} that 
$$
[x_u, +\infty) \subseteq \{x:\log{\overline{F}(x)} \le -\log{p}\} \subseteq [x_l, +\infty),
$$
and so $x_l\le u_p \le x_u$, and the expression for the quantiles follow.
\end{proof}

>>>>

We introduce a fairly general class of \ac{AGG} distributions and \stilian{review their tail behavior,
which will play an important role in the sequel.}{\fbox{check the edit}}

@@

The AGG models include, for example, the standard Gaussian distribution ($\nu = 2$) and the Laplace distribution ($\nu = 1$) as special cases. 
Since the requirement is only placed on the tail behavior, this class encompasses a large variety of light-tailed models. 
This class is commonly used in the literature on high-dimensional testing
 \citep{cai2007estimation, arias2017distribution}.

@@

% \begin{proposition}[$(1 - 1/p)$-th quantile] \label{prop:quantile}
% Let 
% \begin{equation} \label{eq:quantiles-Appendix}
% u_p = F^{\leftarrow}(1-1/p),    
% \end{equation}
% then,
% \begin{equation}
% \lim_{p\to\infty} \frac{u_p}{\left(\nu\log{p}\right)^{1/\nu}} = 1.
% \end{equation}
% \end{proposition}
% 
\begin{proof}[Proof of Proposition \ref{prop:quantile}]
By definition of AGG, for any $\epsilon>0$, there is a constant $C(\epsilon)$ such that for all $x\ge C$, we have
$$
-\frac{1}{\nu}x^\nu(1+\epsilon) \le \log{\overline{F}(x)} \le -\frac{1}{\nu}x^\nu(1-\epsilon).
$$
Therefore, for all $x < x_l := \left((1+\epsilon)^{-1}\nu\log{p}\right)^{1/\nu}$, we have
\begin{equation} \label{eq:AGG-quantiles-proof-1}
    -\log{p} = -\frac{1}{\nu}x_l^\nu(1+\epsilon) \le \log{\overline{F}(x_l)} \le \log{\overline{F}(x)},
\end{equation}
and for all $x > x_u := \left((1-\epsilon)^{-1}\nu\log{p}\right)^{1/\nu}$, we have
\begin{equation} \label{eq:AGG-quantiles-proof-2}
    \log{\overline{F}(x)} \le \log{\overline{F}(x_u)} \le -\frac{1}{\nu}x_u^\nu(1-\epsilon) = -\log{p}.
\end{equation}
By definition of generalized inverse,
\begin{equation*}
    u_p := F^\leftarrow(1-1/p) = \inf\{x:\overline{F}(x)\le 1/p\} = \inf\{x:\log{\overline{F}(x)} \le -\log{p}\}.
\end{equation*}
We know from relations \eqref{eq:AGG-quantiles-proof-1} and \eqref{eq:AGG-quantiles-proof-2} that 
$$
[x_u, +\infty) \subseteq \{x:\log{\overline{F}(x)} \le -\log{p}\} \subseteq [x_l, +\infty),
$$
and so $x_l\le u_p \le x_u$, and the expression for the quantiles follows.
\end{proof}

### 2021-01-08 2.background/2.RV_and_RS.tex
shortened version
* Implemented my corrections up to Ch.3.1

* Moved the proofs of Ch 3 and the sections on non-AGG models from Ch 4 to the Appendix -- which will become the online supplement.

* We need to cut just 5 pages now...  This can happen by re-distributing some proofs in the text (instead of having them in separate sections) and moving some more proofs from Ch.6 to the Appendix.

* Overall, I don't like the idea to eliminate the proofs from Ch. 3.   I also like having the non-AGG models in the main text, too.  I will think if we can shorten the manuscript in alternative ways....

  O

The behavior of the maxima of identically distributed random variables has been studied extensively in the literature (see, e.g., \cite{leadbetter2012extremes,resnick2013extreme,embrechts2013modelling,de2007extreme} 
and the references therein). 
% We show in this subsection that the maxima of errors with rapidly varying tails can be bounded above using quantiles of their marginal distribution, regardless of their dependence structure; this was a key step in the proof of Theorem \ref{thm:sufficient}.
The concept of rapid variation plays an important role in the light-tailed case.

@@

When $F(x)<1$ for all finite $x$, \citet{gnedenko1943distribution} showed that the distribution $F$ has rapidly varying tails if and only if the maxima of independent observations from $F$ are \emph{relatively stable} in the following sense.
\begin{definition}[Relative stability] \label{def:RS}
Let $\epsilon_p = \left(\epsilon_p(i)\right)_{i=1}^p$ be a sequence of random variables with identical marginal distributions $F$. 
Define the sequence $(u_p)_{p=1}^\infty$ to be the $(1-1/p)$-th generalized quantile of $F$, i.e., 
\begin{equation} \label{eq:quantiles}
    u_p = F^\leftarrow(1 - 1/p).
\end{equation}
The triangular array ${\cal E} = \{\epsilon_p, p\in\N\}$ is said to have relatively stable (RS) maxima if
\begin{equation} \label{eq:RS-condition}
    \frac{1}{u_{p}} M_p := \frac{1}{u_{p}} \max_{i=1,\ldots,p} \epsilon_p(i) \xrightarrow{\P} 1,\quad \text{as }\;p\to\infty.
\end{equation}
\end{definition}

In the case of independent and identically distributed $\epsilon_p(i)$'s, \citet{barndorff1963limit} and \citet{resnick1973almost} obtained necessary and sufficient conditions for the \emph{almost sure stability} of maxima, where the convergence in \eqref{eq:RS-condition} holds almost surely.

While relative stability (and almost sure stability) is well-understood in the independent case, the role of dependence has not been fully explored.
We start this investigation with a small refinement of Theorem 2 in \citet{gnedenko1943distribution} valid under arbitrary dependence.

\begin{proposition}[Rapid variation and relative stability] \label{prop:rapid-varying-tails}
Assume that the array ${\cal E}$ consists of identically distributed random 
variables with cumulative distribution function $F$, where $F(x)<1$ for all finite $x>0$. 
\begin{enumerate}
    \item If $F$ has rapidly varying right tail, then for all $\delta>0$,
        \begin{equation} \label{eq:rapid-varying-tails}
            \P\left[\frac{1}{u_p} M_p\le1+\delta\right] \ge 1 - \frac{\overline F((1+\delta)u_p)}{\overline F(u_p)} \to 1.
        \end{equation}
    \item If, in addition, the array ${\cal E}$ has independent entries, then it is relatively stable if and only if $F$ has rapidly varying tail.
    \label{prop:rapid-varying-tails_part-ii}
\end{enumerate}
\end{proposition}

\begin{proof}[Proof of Proposition \ref{prop:rapid-varying-tails}] 
By the union bound and the fact that 
$p\overline F(u_p) \le 1$, we have
\begin{align}\label{e:prop:rapid-varying-tails_part-i-1}
\P [ M_p > (1+\delta)u_p] \le p \overline F((1+\delta)u_p)
 \le \frac{\overline F((1+\delta)u_p)}{\overline F(u_p)}.
\end{align}
In view of \eqref{e:def:rapid-variation} (rapid variation) and the fact that $u_p\to\infty$, as $p\to\infty$, the right-hand side of \eqref{e:prop:rapid-varying-tails_part-i-1} vanishes 
as $p\to\infty$, for all $\delta>0$.  This completes the proof of \eqref{eq:rapid-varying-tails}. Part 2 is a re-statement of the classic result due to Gnedenko in \cite{gnedenko1943distribution}.
\end{proof}

\stilian{ 
\begin{remark} \label{rem:iid-max} Part (1) of Proposition \ref{prop:rapid-varying-tails} is equivalent to 
\begin{equation}\label{e:rem:iid-max}
  \P\left[\frac{1}{u_p} M_p >  1+\delta_p \right] \longrightarrow 0,\ \ \mbox{ as }p\to\infty,
\end{equation}
for some positive sequence $\delta_p\to 0$.  Notice on the other hand that if $M_p^*$ is the maximum of $p$ iid variables with distribution $F$,
the relative stability property entails $M_p^*/u_p \to 1$, in probability, as $p\to\infty$.  Since the sequence $1+\delta_p\to 1$, 
Relation \eqref{e:rem:iid-max} means that the rate of growth of the maxima $M_n$ in ${\cal E}$ cannot be faster than that of the independent
maxima $M_p^*$.  This somewhat surprising fact is true regardless of the dependence structure of ${\cal E}$ and is solely a 
consequence of the rapid variation of $F$.
\end{remark}
}{}

We next demonstrate that Gaussian, Exponential, Laplace, and Gamma distributions all have rapidly varying tails.

@@

More generally, distributions with heavier tails (e.g., log-normal) and lighter tails (e.g., Gompertz) outside the generalized AGG class \eqref{eq:GAGG} may also possess rapidly varying tails;
heavy-tailed distributions like the Pareto and t-distributions, on the other hand, do not.
These alternative classes of models are will be introduced when we study the phase transitions in Chapter \ref{chap:exact-support-recovery}. 
%For brevity, we focus here on the $\text{AGG}(\nu)$ models.

>>>>

The behavior of the maxima of identically distributed random variables has been studied extensively in the extreme 
value theory literature \citep[see, e.g.,][and the references therein]{leadbetter2012extremes,resnick2013extreme,embrechts2013modelling,de2007extreme}. 
% We show in this subsection that the maxima of errors with rapidly varying tails can be bounded above using quantiles of their marginal distribution, regardless of their dependence structure; this was a key step in the proof of Theorem \ref{thm:sufficient}.
The concept of rapid variation plays an important role in the light-tailed case.

@@

When $F(x)<1$ for all finite $x$, \citet{gnedenko1943distribution} showed that the distribution $F$ has rapidly varying tails if and only if the maxima of independent observations from $F$ are \emph{relatively stable} in the following sense.
\begin{definition}[Relative stability] \label{def:RS}
Let $\epsilon_p = \left(\epsilon_p(i)\right)_{i=1}^p$ be a sequence of random variables with common marginal 
distribution $F$. Define the sequence $(u_p)_{p=1}^\infty$ to be the $(1-1/p)$-th generalized quantile of $F$, i.e., 
\begin{equation} \label{eq:quantiles}
    u_p = F^\leftarrow(1 - 1/p).
\end{equation}
The triangular array ${\cal E} = \{\epsilon_p, p\in\N\}$ is said to have relatively stable (RS) maxima if
\begin{equation} \label{eq:RS-condition}
    \frac{1}{u_{p}} M_p := \frac{1}{u_{p}} \max_{i=1,\ldots,p} \epsilon_p(i) \xrightarrow{\P} 1,\quad \text{as }\;p\to\infty.
\end{equation}
\end{definition}

In the case of independent and identically distributed $\epsilon_p(i)$'s, \citet{barndorff1963limit} and \citet{resnick1973almost} obtained necessary and sufficient conditions for the \emph{almost sure stability} of maxima, where the convergence in \eqref{eq:RS-condition} holds almost surely. \stilian{See also \cite{klass:1984} for further 
sharp results on almost sure stability.  Here, we will only need the weaker notion in \eqref{eq:RS-condition} 
but extend our inquiry to the case of dependent $\epsilon_p(i)$'s.}{\fbox{The ref suggested by a reviewer -- Resnick?}}

While relative stability (and almost sure stability) is well-understood in the independent case, the role of dependence has not been fully explored.
We start this investigation with a small refinement of Theorem 2 in \citet{gnedenko1943distribution} valid under {\em
arbitrary dependence.}

\begin{proposition}[Rapid variation and relative stability] \label{prop:rapid-varying-tails}
Assume that the array ${\cal E}$ consists of identically distributed and possibly dependent random 
variables with cumulative distribution function $F$, where $F(x)<1$ for all finite $x>0$. 
\begin{enumerate}
    \item If $F$ has rapidly varying right tail in the sense of \eqref{e:def:rapid-variation}, then for all $\delta>0$,
        \begin{equation} \label{eq:rapid-varying-tails}
            \P\left[\frac{1}{u_p} M_p\le1+\delta\right] \ge 1 - \frac{\overline F((1+\delta)u_p)}{\overline F(u_p)} \to 1.
        \end{equation}
    \item If the array ${\cal E}$ has independent entries, then it is relatively stable if and only if $F$ has rapidly varying tail, i.e., \eqref{def:rapid-variation} holds.
    \label{prop:rapid-varying-tails_part-ii}
\end{enumerate}
\end{proposition}

\begin{proof}[Proof of Proposition \ref{prop:rapid-varying-tails}] 
By the union bound and the fact that 
$p\overline F(u_p) \le 1$, we have
\begin{align}\label{e:prop:rapid-varying-tails_part-i-1}
\P [ M_p > (1+\delta)u_p] \le p \overline F((1+\delta)u_p)
 \le \frac{\overline F((1+\delta)u_p)}{\overline F(u_p)}.
\end{align}
In view of \eqref{e:def:rapid-variation} (rapid variation) and the fact that $u_p\to\infty$, as $p\to\infty$, the right-hand side of \eqref{e:prop:rapid-varying-tails_part-i-1} vanishes 
as $p\to\infty$, for all $\delta>0$.  This completes the proof of \eqref{eq:rapid-varying-tails}. Part 2 is a re-statement of a classic result dating back to \cite{gnedenko1943distribution}.
\end{proof}

\stilian{ 
\begin{remark} \label{rem:iid-max} Part (1) of Proposition \ref{prop:rapid-varying-tails} is equivalent to 
\begin{equation}\label{e:rem:iid-max}
  \P\left[\frac{1}{u_p} M_p >  1+\delta_p \right] \longrightarrow 0,\ \ \mbox{ as }p\to\infty,
\end{equation}
for some positive sequence $\delta_p\to 0$.  Notice on the other hand that if $M_p^*$ is the maximum of $p$ iid variables with distribution $F$,
the relative stability property entails $M_p^*/u_p \to 1$, in probability, as $p\to\infty$.  Since the sequence $1+\delta_p\to 1$, 
Relation \eqref{e:rem:iid-max} means that the rate of growth of the maxima $M_n$ in ${\cal E}$ cannot be faster than that of the independent
maxima $M_p^*$.  This somewhat surprising fact is true regardless of the dependence structure of ${\cal E}$ and is solely a 
consequence of the rapid variation of $F$.
\end{remark}
}{\fbox{The above remark addresses a question of a reviewer}}

We demonstrate next that the Gaussian, Exponential, Laplace, and Gamma distributions all have rapidly varying tails.

@@

More generally, distributions with heavier tails (e.g., log-normal) and lighter tails (e.g., Gompertz) outside the generalized AGG class \eqref{eq:GAGG} may also possess rapidly varying tails;
heavy-tailed distributions like the Pareto and t-distributions, on the other hand, do not.
\stilian{More details on these alternative classes of models can be found in the Supplement.}
{\fbox{Move these to the online Supplement?} 
These alternative classes of models will be introduced when we study the phase transitions in 
Chapter \ref{chap:exact-support-recovery}.}
%For brevity, we focus here on the $\text{AGG}(\nu)$ models.
